Investigating a dataset by Bidish Basu
========================================================
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r ,message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr)
library(dplyr)
library(gridExtra)

```

```{r Load_the_Data}
# Load the Data
df<-read.csv("wineQualityReds.csv")
#Printing out a few columns
head(df)
```

In this project, I will be exploring the "wineQualityReds.csv" dataset.

# Univariate Plots Section



```{r  inital_analysis1}
#Exploring the dataset
summary(df)
```
```{r  inital_analysis2}
#Removing column X
df$X<-NULL
#Printing the dataframe
head(df)
```
```{r  inital_analysis3}
str(df)
```

Here are my observations from the prelimnary analysis,

1. There are 1599 rows and 13 columns.
2. I have removed the column named "X" as it was not of much importance to 
     my analysis after which I am left with 1599 rows and 12 columns.
3. As per the information given about the dataset, the variable "quality"
      catches my eye, as it is a categorical variable which will help me in 
      my analysis.
  

 Since "quality" is my main focus here, I will be creating a new variable to
  further simplify the way the quality of the wine is assessed. I am going to 
  divide the dataset based  on the quality rating,as follows

 1. If the quality is rated between 0 to 4 then its "low"
  2. If the quality is rated between 5 to 7 then its "medium"
  3. If the quality is rated above 7 then its "high"
  
```{r  New_Variable}
#Creating the new column

df$wine_rating<-ifelse(df$quality <= 4,"low",ifelse((df$quality >= 5) & 
(df$quality <= 7), "medium", "high"))

#Printing some rows
head(df)
```


We see that a new column has been added to the dataset called "wine_rating" 
where the wines are rated as "high","medium" and "low" based on their quality.

I am converting the "quality" and "wine_rating" variables to factors.

```{r  inital_analysis4}
#Converting "quality" and "wine_rating" to a factor
df$quality<-factor(df$quality)
df$wine_rating<-factor(df$wine_rating)
str(df)
```
```{r  inital_analysis5}
table(df$quality)
```

We see that quality has been converted to a factor with 6 levels ranging from 
3-8

```{r  inital_analysis6}
table(df$wine_rating)
```

We see that wine_rating has been converted to a factor with 3 levels.

```{r message=FALSE, warning=FALSE,Univariate_Plots1}

p1<- ggplot(data=df,aes(x=fixed.acidity))+geom_histogram(fill='red')
p2<- ggplot(data=df,aes(x=volatile.acidity))+geom_histogram(fill='blue')
p3<- ggplot(data=df,aes(x=citric.acid))+geom_histogram(fill='green')
p4<- ggplot(data=df,aes(x=residual.sugar))+geom_histogram(fill='blue')+
  scale_x_log10()
p5<- ggplot(data=df,aes(x=chlorides))+geom_histogram(fill='green')+
  scale_x_log10()
p6<- ggplot(data=df,aes(x=free.sulfur.dioxide))+geom_histogram(fill='red')
p7<- ggplot(data=df,aes(x=total.sulfur.dioxide))+geom_histogram(fill='green')
p8<- ggplot(data=df,aes(x=density))+geom_histogram(fill='red')
p9<- ggplot(data=df,aes(x=pH))+geom_histogram(fill='blue')
p10<- ggplot(data=df,aes(x=sulphates))+geom_histogram(fill='red')
p11<- ggplot(data=df,aes(x=alcohol))+geom_histogram(fill='green')

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol=4)
```

 We see from the histograms that, 
   1. Density and pH are normally distributed.
  2. Fixed and volatile acidity,sulphates,alcohol,chlorides,residual sugar
       are long tailed.  


```{r  Univariate_Plots3}
ggplot(data=df,aes(x=wine_rating))+geom_bar(aes(fill=wine_rating))

```

 We see that as per the ratings, we have a large amount of wine that have quality 
  between 5-7 as compared to ones below 5 and above 7 which are quite small in 
  number.



# Univariate Analysis


### What is the structure of your dataset?

There are 1599 observations in the dataset with 13 variables.

The 13 features and their short description is as follows:

Fixed acidity: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)

Volatile acidity: the amount of acetic acid in wine.

Citric acid: mostly used for adding flavor.

Residual sugar: the amount of sugar remaining after fermentation stops 
and measured in gms of sugar per liter.

Chlorides: the amount of salt in the wine.

Free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine.

Total sulfur dioxide: amount of free and bound forms of S02.

Density: the density of water is close to that of water depending on the 
percent alcohol and sugar content.

pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) 
to 14 (very basic); most wines are between 3-4 on the pH scale.

Sulphates: a wine additive which can contribute to sulfur dioxide gas (S02)
levels, which acts as an antimicrobial and antioxidant.

Alcohol: the percent alcohol present in the wine.

Quality: rating of the wine between 0-10 in the given sample

wine_rating: rating of wine as high,medium,low.

### What is/are the main feature(s) of interest in your dataset?

The main feature that intrests me is the quality of the wine.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Features like fixed and volatile acidity, sulfur dioxide, alcohol and pH 
  are interesting as well.

### Did you create any new variables from existing variables in the dataset?

I created a new variable called "wine_rating" depending on the "quality" of 
wine.
### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I found citic acid to be very unusually distributed. But since that is not of
  importance to my investigation, I have not tidied it up.


# Bivariate Plots Section



```{r  Bivariate_Plots}

p1<- ggplot(data=df,aes(y=fixed.acidity,x=wine_rating))+geom_boxplot()
p2<- ggplot(data=df,aes(y=volatile.acidity,x=wine_rating))+geom_boxplot()
p3<- ggplot(data=df,aes(y=citric.acid,x=wine_rating))+geom_boxplot()
p4<- ggplot(data=df,aes(y=residual.sugar,x=wine_rating))+geom_boxplot()
p5<- ggplot(data=df,aes(y=chlorides,x=wine_rating))+geom_boxplot()
p6<- ggplot(data=df,aes(y=free.sulfur.dioxide,x=wine_rating))+geom_boxplot()
p7<- ggplot(data=df,aes(y=total.sulfur.dioxide,x=wine_rating))+geom_boxplot()
p8<- ggplot(data=df,aes(y=density,x=wine_rating))+geom_boxplot()
p9<- ggplot(data=df,aes(y=pH,x=wine_rating))+geom_boxplot()
p10<- ggplot(data=df,aes(y=sulphates,x=wine_rating))+geom_boxplot()
p11<- ggplot(data=df,aes(y=alcohol,x=wine_rating))+geom_boxplot()


grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol=4)

```

From the above plot, we see that the wines that are rated "medium" have alot 
of outliers across all the variables. Let's look further into this.

```{r  Bivariate_Plots1}
p1<- ggplot(data=df,aes(y=fixed.acidity,x=quality))+geom_boxplot()
p2<- ggplot(data=df,aes(y=volatile.acidity,x=quality))+geom_boxplot()
p3<- ggplot(data=df,aes(y=citric.acid,x=quality))+geom_boxplot()
p4<- ggplot(data=df,aes(y=residual.sugar,x=quality))+geom_boxplot()
p5<- ggplot(data=df,aes(y=chlorides,x=quality))+geom_boxplot()
p6<- ggplot(data=df,aes(y=free.sulfur.dioxide,x=quality))+geom_boxplot()
p7<- ggplot(data=df,aes(y=total.sulfur.dioxide,x=quality))+geom_boxplot()
p8<- ggplot(data=df,aes(y=density,x=quality))+geom_boxplot()
p9<- ggplot(data=df,aes(y=pH,x=quality))+geom_boxplot()
p10<- ggplot(data=df,aes(y=sulphates,x=quality))+geom_boxplot()
p11<- ggplot(data=df,aes(y=alcohol,x=quality))+geom_boxplot()


grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol=4)
```

When we plot a similar plot using "quality", the wines that are rated between 
4-7 have  high outliers.


```{r  Bivariate_Plots3}

p1<-ggplot(data=df,aes(x=citric.acid,y=fixed.acidity))+
  geom_jitter(alpha=0.25,color='cyan')+geom_smooth(method='lm',colour='blue')

p2<-ggplot(data=df,aes(x=density,y=fixed.acidity))+
  geom_jitter(alpha=0.25,color='green')+geom_smooth(method='lm',colour='red')

p3<-ggplot(data=df,aes(x=pH,y=fixed.acidity))+
  geom_jitter(alpha=0.25,color='purple')+geom_smooth(method='lm',colour='blue')

p4<-ggplot(data=df,aes(x=citric.acid,y=volatile.acidity))+
  geom_jitter(alpha=0.25,color='yellow')+geom_smooth(method='lm',colour='red')

grid.arrange(p1,p2,p3,p4,ncol=2)


```

We see that, there are strong positive correlations between fixed acidity 
and density and  citric acid.
  
Similarly, we see that there is a strong negative correlation between fixed
acidity and pH  also with volatile acid and ctric acid.

```{r  Bivariate_Plots4}

  p1<- ggplot(data = df,aes(x=quality,y=alcohol))+
  geom_jitter(color='purple')+
       geom_boxplot(alpha=0.5,color='blue')
  p2<- ggplot(data = df,aes(x=quality,y=volatile.acidity))+
    geom_jitter(color='cyan')+
       geom_boxplot(alpha=0.5,color='green')
  p3<- ggplot(data = df,aes(x=quality,y=citric.acid))+
    geom_jitter(color='yellow')+
       geom_boxplot(alpha=0.5,color='red')
  
  grid.arrange(p1,p2,p3,nrow=1)
```

From the box plots above, we see that, alcohol and ctric acid have strong 
positive correlations with quality and volatile acid has a strong negative 
correlation.


```{r  Bivariate_Plots6}

wine_subset <- df[ , !names(df) %in% c("wine_rating")]
cor(wine_subset[, c(1:11)])
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Scatterplots and boxplots showed strong positive correlation of alcohol with 
quality and strong negative correlation between volatile acidity and quality.

So, we infer that good wine generally has higher alcohol content,higher 
citric acidity and lower volatile acidity.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

We see some strong correlations as follows,
  

Fixed acidity vs citric acid (0.67)

Volatile acidity vs citric acid (-0.55)

Fixed acidity vs density (0.67)

Fixed acidity vs pH (-0.68)

### What was the strongest relationship you found?

Citric acid and fixed acidity showed a strong positive correlation (68%)

pH and fixed acidity showed a strong negative correlation(-68%)

# Multivariate Plots Section


```{r  Multivariate_Plots}

library(GGally)
df_sub<-df[,c(1,2,3,8,9,13)]
ggpairs(df_sub, aes(colour = wine_rating, alpha = 0.6))


```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Good quality wines tend to have higher citric acid and lower volatile acidity
levels.
To predict the quality of the wines we can use Alcohol, citric acid,
volatile acidity,sulphates to create a linear model.
pH according to my my reserach plays a big role in determinig the quality of
wine. But the sample given does not seem to be indicative of that.

### Were there any interesting or surprising interactions between features?


One of the surprising aspects of my analysis is that the pH did not really have 
a significant role to play in determining wine quality. As such most good wines 
have pH between 3-4.


### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

As I have stated in the previous section, pH should have played an important 
role in determinig the quality of the wines but the sample data is not 
indicative of the same. From the previous sections we see that the pH levels 
as per the box plots generated are almost the same for the highand medium 
quality wines.


------

# Final Plots and Summary


### Plot One
```{r Plot_One}

ggplot(data = df,aes(x=fixed.acidity,y=citric.acid))+
  geom_point(aes(color=wine_rating),alpha = 0.4)+
  ggtitle("Citric Acid and Fixed acidity")+xlab('Fixed Acidity ( g / dm^3)')+
  ylab('Citric Acid (g / dm^3)')
```

### Description One

 Both Citric Acid and Fixed Acidity have a big role in determinig the quality 
of wine.
From the plot above we see that, high quality wines have high levels of the 
both.

### Plot Two
```{r  Plot_Two}

ggplot(data = df,aes(x=alcohol,y=density))+
  geom_point(aes(color=wine_rating),alpha = 0.4)+
  ggtitle("Alcohol and Density")+xlab('Alcohol (% by volume)')+ylab('Density (g / cm^3)')

```

### Description Two
Alcohol and density also play an important role in determinig the quality 
of the wine. From the plot we see that higher the quality of wine, higher is 
the alcohol and lower is the density.

### Plot Three
```{r  Plot_Three}
ggplot(data = df,aes(x=alcohol,y=residual.sugar))+
  geom_point(aes(color=wine_rating),alpha = 0.4)+
  ggtitle("Alcohol and Residual Sugar")+xlab('Alcohol (% by volume)')+
  ylab('Residual Sugar (g / dm^3)')

```

### Description Three
 From the plot above, we see that, high quality wines have low amounts 
of residual sugar and high amounts of alcohol.

------

# Reflection

In this analysis, my main struggle was to find the factors that are 
responsible for the production of different quality of wines especially the 
'high' and the 'low' ones. As the data was very centralized towards the 
'medium' quality. So maybe in future, I can get a dataset about Red Wines with
more complete information so that I can build my models more effectively.
I was happy to see that the 'high' quality wines have high amounts of alcohol 
and low amounts of residual sugar and volatile acids.
For future analysis, I would love to have a dataset, where a rank is given for
that particular wine by a number of different wine tasters. This is because, 
when we include the human element, we get multiple opinions so by including
the human element in my analysis, I would be able to have a better understanding
of the quality of wine. Having these factors included inside the dataset would
result in a different insight altogether in my analysis.

